Stat 534 Project: \\ Extrapolation from Poisson Process Intensity Surface Models

### 2017-03-23 project/paper.rnw
Done with the presentation.

<<polyfit1, warning = FALSE, fig.width = 6, fig.height = 8, fig.pos = 'p', fig.cap = ''>>=
# Create a quadrature scheme manually to ensure there are dummy points on
# each transect.
# TODO: Figure out weights.
ex_Q <- quad(data = ex_ppp,
             dummy = as.ppp(rbind(
               data.frame(vertices(Window(ex_ppp))),
               do.call(rbind, lapply(unique(ex_path$x), function(x){
                 return(data.frame(
                   x = x, y = seq(min(ex_path$y), max(ex_path$y),
                                  length.out = 128)
                 ))}))), W = Window(ex_ppp)))

ex_fit2 <- ppm(ex_Q, ~ polynom(scale(x, center = 1556870, scale = 7625),
                               scale(y, center = 535421, scale = 5709),
                                     2), correction = 'iso')
ex_fit3 <- ppm(ex_Q, ~ polynom(scale(x, center = 1556870, scale = 7625),
                               scale(y, center = 535421, scale = 5709),
                                     3), correction = 'iso')
ex_fit4 <- ppm(ex_Q, ~ polynom(scale(x, center = 1556870, scale = 7625),
                               scale(y, center = 535421, scale = 5709),
                                     4), correction = 'iso')
ex_fit5 <- ppm(ex_Q, ~ polynom(scale(x, center = 1556870, scale = 7625),
                               scale(y, center = 535421, scale = 5709),
                                     5), correction = 'iso')
ex_fit6 <- ppm(ex_Q, ~ polynom(scale(x, center = 1556870, scale = 7625),
                               scale(y, center = 535421, scale = 5709),
                                     6), correction = 'iso')
ex_fit7 <- ppm(ex_Q, ~ polynom(scale(x, center = 1556870, scale = 7625),
                               scale(y, center = 535421, scale = 5709),
                                     7), correction = 'iso')
ex_fit8 <- ppm(ex_Q, ~ polynom(scale(x, center = 1556870, scale = 7625),
                               scale(y, center = 535421, scale = 5709),
                                     8), correction = 'iso')
ex_fit9 <- ppm(ex_Q, ~ polynom(scale(x, center = 1556870, scale = 7625),
                               scale(y, center = 535421, scale = 5709),
                                     9), correction = 'iso')
ex_fit10 <- ppm(ex_Q, ~ polynom(scale(x, center = 1556870, scale = 7625),
                                scale(y, center = 535421, scale = 5709),
                                      10), correction = 'iso')
ex_fit11 <- ppm(ex_Q, ~ polynom(scale(x, center = 1556870, scale = 7625),
                                scale(y, center = 535421, scale = 5709),
                                      11), correction = 'iso')
ex_fit12 <- ppm(ex_Q, ~ polynom(scale(x, center = 1556870, scale = 7625),
                                scale(y, center = 535421, scale = 5709),
                                      12), correction = 'iso')

@@

# Multiply predictions by 43560 to convert from anomalies per square foot
# to anomalies per acre.
plot(43560 * predict(ex_fit2, window = site_window), main = '2nd Degree')
plot(43560 * predict(ex_fit3, window = site_window), main = '3rd Degree')
plot(43560 * predict(ex_fit4, window = site_window), main = '4th Degree')
plot(43560 * predict(ex_fit5, window = site_window), main = '5th Degree')
plot(43560 * predict(ex_fit6, window = site_window), main = '6th Degree')
plot(43560 * predict(ex_fit7, window = site_window), main = '7th Degree')
plot(43560 * predict(ex_fit8, window = site_window), main = '8th Degree')
plot(43560 * predict(ex_fit9, window = site_window), main = '9th Degree')
plot(43560 * predict(ex_fit10, window = site_window), main = '10th Degree')
plot(43560 * predict(ex_fit11, window = site_window), main = '11th Degree')
plot(43560 * predict(ex_fit12, window = site_window), main = '12th Degree')
plot(43560 * density(ex_ppp_full) * area(ex_ppp_full) / ex_ppp_full$n,
     main = 'Kernel Density')
@

\texttt{plot} doesn't work, and \texttt{predict} doesn't work on the original
window. \texttt{predict} can't compute SEs.

>>>>

<<polyfit1, cache = TRUE, warning = FALSE, fig.width = 6, fig.height = 8, fig.pos = 'p', fig.cap = ''>>=
# Create a quadrature scheme manually to ensure there are dummy points on
# each transect.
# TODO: Figure out weights.
ex_Q <- quad(data = ex_ppp,
             dummy = as.ppp(rbind(
               data.frame(vertices(Window(ex_ppp))),
               do.call(rbind, lapply(unique(ex_path$x), function(x){
                 return(data.frame(
                   x = x, y = seq(min(ex_path$y), max(ex_path$y),
                                  length.out = 128)
                 ))}))), W = Window(ex_ppp)))

ex_fit2 <- ppm(ex_Q, ~ polynom(scale(x), scale(y), 2), correction = 'iso')
ex_fit3 <- ppm(ex_Q, ~ polynom(scale(x), scale(y), 3), correction = 'iso')
ex_fit4 <- ppm(ex_Q, ~ polynom(scale(x), scale(y), 4), correction = 'iso')
ex_fit5 <- ppm(ex_Q, ~ polynom(scale(x), scale(y), 5), correction = 'iso')
ex_fit6 <- ppm(ex_Q, ~ polynom(scale(x), scale(y), 6), correction = 'iso')
ex_fit7 <- ppm(ex_Q, ~ polynom(scale(x), scale(y), 7), correction = 'iso')
ex_fit8 <- ppm(ex_Q, ~ polynom(scale(x), scale(y), 8), correction = 'iso')
ex_fit9 <- ppm(ex_Q, ~ polynom(scale(x), scale(y), 9), correction = 'iso')
ex_fit10 <- ppm(ex_Q, ~ polynom(scale(x), scale(y), 10), correction = 'iso')
ex_fit11 <- ppm(ex_Q, ~ polynom(scale(x), scale(y), 11), correction = 'iso')
ex_fit12 <- ppm(ex_Q, ~ polynom(scale(x), scale(y), 12), correction = 'iso')

# Get predictions and SEs.
ex_p2 <- predict(ex_fit2, window = site_window, se = TRUE)
ex_p3 <- predict(ex_fit3, window = site_window, se = TRUE)
ex_p4 <- predict(ex_fit4, window = site_window, se = TRUE)
ex_p5 <- predict(ex_fit5, window = site_window, se = TRUE)
ex_p6 <- predict(ex_fit6, window = site_window, se = TRUE)
ex_p7 <- predict(ex_fit7, window = site_window, se = TRUE)
ex_p8 <- predict(ex_fit8, window = site_window, se = TRUE)
ex_p9 <- predict(ex_fit9, window = site_window, se = TRUE)
ex_p10 <- predict(ex_fit10, window = site_window, se = TRUE)
ex_p11 <- predict(ex_fit11, window = site_window, se = TRUE)
ex_p12 <- predict(ex_fit12, window = site_window, se = TRUE)

@@

# Multiply predictions by 43560 to convert from anomalies per square foot
# to anomalies per acre.
plot(43560 * ex_p2$estimate, main = '2nd Degree')
plot(43560 * ex_p3$estimate, main = '3rd Degree')
plot(43560 * ex_p4$estimate, main = '4th Degree')
plot(43560 * ex_p5$estimate, main = '5th Degree')
plot(43560 * ex_p6$estimate, main = '6th Degree')
plot(43560 * ex_p7$estimate, main = '7th Degree')
plot(43560 * ex_p8$estimate, main = '8th Degree')
plot(43560 * ex_p9$estimate, main = '9th Degree')
plot(43560 * ex_p10$estimate, main = '10th Degree')
plot(43560 * ex_p11$estimate, main = '11th Degree')
plot(43560 * ex_p12$estimate, main = '12th Degree')
plot(43560 * density(ex_ppp_full) * area(ex_ppp_full) / ex_ppp_full$n,
     main = 'Kernel Density')
@

<<polyse1, fig.width = 6, fig.height = 8, fig.pos = 'p', fig.cap = ''>>=
par(mfrow = c(4, 3), mar = c(1, 0, 2, 3), las = 2, cex = 1)
plot(log(43560 * ex_p2$se), main = '2nd Degree\nlog(SE)')
plot(log(43560 * ex_p4$se), main = '4th Degree\nlog(SE)')
plot(log(43560 * ex_p6$se), main = '6th Degree\nlog(SE)')
plot(ex_p2$se / ex_p2$estimate, main = '2nd Degree\nRSE')
plot(ex_p4$se / ex_p4$estimate, main = '4th Degree\nRSE')
plot(ex_p6$se / ex_p6$estimate, main = '6th Degree\nRSE')

plot(log(43560 * ex_p8$se), main = '8th Degree\nlog(SE)')
plot(log(43560 * ex_p10$se), main = '10th Degree\nlog(SE)')
plot(log(43560 * ex_p12$se), main = '12th Degree\nlog(SE)')
plot(ex_p8$se / ex_p8$estimate, main = '8th Degree\nRSE')
plot(ex_p10$se / ex_p10$estimate, main = '10th Degree\nRSE')
plot(ex_p12$se / ex_p12$estimate, main = '12th Degree\nRSE')
@

\texttt{plot} doesn't work, and \texttt{predict} doesn't work on the original
window

### 2017-03-23 project/pres.rnw
Done with the presentation.

\item Mapping geomegnetic anomalies prior to an unexploded ordnace (UXO)
remediation

@@

\item Their method is implemented in \texttt{spatstat}'s \texttt{ppm}, with
\texttt{glm} from base R or \texttt{gam} from \texttt{mgcv} as the back-end

@@

<<uxoexampledata>>=
# Define the window for the UXO site.
site_window <- owin(poly = cbind(x = c(1564294, 1564495, 1556870, 1557126),
                                 y = c(535421, 541130, 541085, 535576)),
                    unitname = c('foot', 'feet'))
@

@@

>>>>

\item Mapping geomagnetic anomalies prior to an unexploded ordnance (UXO)
remediation

@@

\item Their method is implemented in \texttt{spatstat}'s \texttt{ppm}, with
\texttt{glm} from base R or \texttt{gam} from \texttt{mgcv} as the back-end,
using the \texttt{quasi} family

@@

<<uxoexampledata>>=
site_window <- owin(poly = cbind(x = c(1564294, 1564495, 1556870, 1557126),
                                 y = c(535421, 541130, 541085, 535576)),
                    unitname = c('foot', 'feet'))
ex_anom <- read.csv('easy_sample_tTA2_p200_bg100_fg200_rep2000.anomaly')
ex_path <- read.csv('easy_sample_tTA2_p200_bg100_fg200_rep2000.cog')
ex_win <- intersect.owin(site_window,
  do.call(union.owin, lapply(unique(ex_path$x), function(x){
    return(owin(c(x - 3, x + 3), c(535421, 541130),
           unitname = c('foot', 'feet')))
  })))
ex_ppp <- ppp(ex_anom$x, ex_anom$y, window = ex_win)
ex_ppp_full <- ppp(ex_anom$x, ex_anom$y, window = site_window)
@

@@

\begin{frame}
\frametitle{Observed Events}

\item Metal detectors record anomalies in six foot wide strips along parallel
transects with 396 feet between centerlines

\item Observed \Sexpr{sprintf('%.1f', area(ex_win) / 43560)} acres,
\Sexpr{sprintf('%.1f\\%%', 100 * area(ex_win) / area(site_window))}
of the site

<<sim2000obs>>=
par(mar = c(0, 0, 1, 0))
plot(ex_ppp, main = 'Observed Geomagnetic Anomalies', pch = '.', border = NA)
plot(site_window, add = TRUE)
@
\end{frame}

\begin{frame}
\frametitle{Trend Surface Models}

\item Polynomial models
\begin{equation*}
\log(\lambda(x,y)) = \sum_{i=0}^{p}\sum_{j=0}^{p-i}\beta_{ij}x^{i}y^{j};
\qquad p = 2, 3, \dots, 12
\end{equation*}

\item Can approximate complicated surfaces

\item Expect two peaks, so even \(p \geq 4\) could work well

\item Rescaling \(x\) and \(y\) to mean 0 and variance 1 reduces numerical
instability for large \(p\)

\begin{frame}
\frametitle{Extrapolated Surfaces}

<<polyfit, warning = FALSE, cache = TRUE>>=
ex_fit1 <- ppm(ex_ppp ~ x + y)
ex_Q <- quad(data = ex_ppp,
             dummy = as.ppp(rbind(
               data.frame(vertices(Window(ex_ppp))),
               do.call(rbind, lapply(unique(ex_path$x), function(x){
                 return(data.frame(
                   x = x, y = seq(min(ex_path$y), max(ex_path$y),
                                  length.out = 128)
                 ))}))), W = Window(ex_ppp)))

ex_fit2 <- ppm(ex_Q, ~ polynom(scale(x), scale(y), 2), correction = 'iso')
ex_fit3 <- ppm(ex_Q, ~ polynom(scale(x), scale(y), 3), correction = 'iso')
ex_fit4 <- ppm(ex_Q, ~ polynom(scale(x), scale(y), 4), correction = 'iso')
ex_fit5 <- ppm(ex_Q, ~ polynom(scale(x), scale(y), 5), correction = 'iso')
ex_fit6 <- ppm(ex_Q, ~ polynom(scale(x), scale(y), 6), correction = 'iso')
ex_fit7 <- ppm(ex_Q, ~ polynom(scale(x), scale(y), 7), correction = 'iso')
ex_fit8 <- ppm(ex_Q, ~ polynom(scale(x), scale(y), 8), correction = 'iso')
ex_fit9 <- ppm(ex_Q, ~ polynom(scale(x), scale(y), 9), correction = 'iso')
ex_fit10 <- ppm(ex_Q, ~ polynom(scale(x), scale(y), 10), correction = 'iso')
ex_fit11 <- ppm(ex_Q, ~ polynom(scale(x), scale(y), 11), correction = 'iso')
ex_fit12 <- ppm(ex_Q, ~ polynom(scale(x), scale(y), 12), correction = 'iso')

ex_p2 <- predict(ex_fit2, window = site_window, se = TRUE)
ex_p3 <- predict(ex_fit3, window = site_window, se = TRUE)
ex_p4 <- predict(ex_fit4, window = site_window, se = TRUE)
ex_p5 <- predict(ex_fit5, window = site_window, se = TRUE)
ex_p6 <- predict(ex_fit6, window = site_window, se = TRUE)
ex_p7 <- predict(ex_fit7, window = site_window, se = TRUE)
ex_p8 <- predict(ex_fit8, window = site_window, se = TRUE)
ex_p9 <- predict(ex_fit9, window = site_window, se = TRUE)
ex_p10 <- predict(ex_fit10, window = site_window, se = TRUE)
ex_p11 <- predict(ex_fit11, window = site_window, se = TRUE)
ex_p12 <- predict(ex_fit12, window = site_window, se = TRUE)
@

<<resplots1>>=
par(mfrow = c(2, 3), mar = c(1, 0, 1.2, 3), las = 2, cex = 1)
plot(43560 * ex_p2$estimate, main = '2nd Degree Polynom')
plot(43560 * ex_p3$estimate, main = '3rd Degree Polynom')
plot(43560 * ex_p4$estimate, main = '4th Degree Polynom')
plot(43560 * ex_p5$estimate, main = '5th Degree Polynom')
plot(43560 * ex_p6$estimate, main = '6th Degree Polynom')
plot(43560 * ex_p7$estimate, main = '7th Degree Polynom')
@

\begin{frame}
\frametitle{Extrapolated Surfaces}

<<resplots2>>=
par(mfrow = c(2, 3), mar = c(1, 0, 1.2, 3), las = 2, cex = 1)
plot(43560 * ex_p8$estimate, main = '8th Degree Polynom')
plot(43560 * ex_p9$estimate, main = '9th Degree Polynom')
plot(43560 * ex_p10$estimate, main = '10th Degree Polynom')
plot(43560 * ex_p11$estimate, main = '11th Degree Polynom')
plot(43560 * ex_p12$estimate, main = '12th Degree Polynom')
plot(43560 * density(ex_ppp_full) * area(ex_ppp_full) / area(ex_ppp),
     main = 'Kernel Density')
@

\begin{frame}
\frametitle{Standard Errors}

<<seplots>>=
par(mfrow = c(2, 3), mar = c(1, 0, 1.2, 3), las = 2, cex = 1)
plot(log(43560 * ex_p2$se), main = '2nd Degree log(SE)')
plot(log(43560 * ex_p4$se), main = '4th Degree log(SE)')
plot(log(43560 * ex_p6$se), main = '6th Degree log(SE)')
plot(log(43560 * ex_p8$se), main = '8th Degree log(SE)')
plot(log(43560 * ex_p10$se), main = '10th Degree log(SE)')
plot(log(43560 * ex_p12$se), main = '12th Degree log(SE)')
@

\begin{frame}
\frametitle{Problems with Implementation}

\item By default, \texttt{ppm} places dummy points on a grid across a bounding
box

\item Dummy points outside the actual window are discarded

\item Only \Sexpr{prettyNum(ex_fit1$Q$dummy$n, big.mark = ',')} points on two
transects are kept

\item I place 128 evenly-spaced dummy points along each transect

\item \Sexpr{prettyNum(ex_Q$dummy$n, big.mark = ',')} are used

<<dummy>>=
par(mfrow = c(1, 2), mar = c(1, 0, 1, 0), cex = 1)
plot(ex_fit1$Q$dummy, pch = '.', border = NA,
     main = 'Default Dummy Points')
plot(site_window, add = TRUE)
plot(ex_Q$dummy, pch = '.', border = NA,
     main = 'Manual Dummy Points')
plot(site_window, add = TRUE)
@

\begin{frame}
\frametitle{Problems with Implementation}

\item For \(p \geq 18\), \texttt{ppm} cannot compute SEs because the
``Fisher information matrix is singular'' --- 190 coefficients

\item The \texttt{plot} method gives an error about infinite values

\item When the window isn't specified, the \texttt{predict} method's default
grid misses all but two transects

\item The predict method does not work with spline smoothers

\item The magnitudes of the predictions are much too large

\begin{frame}
\frametitle{Conclusion}

\item Polynomial surfaces are flexible but require much faith in the model

\item How to check these models?

\item How much of the region must be observed?

\item Can implement with existing R packages but not easily

\item What scale are the prediction SEs on?
